---
title: "UNICEF Data Visualisations"
subtitle: "Exploring Population and Life Expectancy Growth from 1960 to 2021"
author: "Sean Hurley"
date: 10-Apr-2024
format: 
  html:
    theme: lux
    toc: true
    toc-title: Content Outline
    fig-align: left
  pdf: default
editor: visual
execute:
  echo: false
---



```{r}
#| label: Install Packages
#| echo: false
#| output: false
if(!require("tidyverse")) install.packages("tidyverse", dependencies = TRUE)
```

```{r}
#| label: Load Packages
#| echo: false
#| output: false
library("tidyverse")
library("ggplot2")
```

```{r}
#| echo: false
#| output: false
getwd() # function to check the current working directory
```

```{r}
#| echo: false
#| output: false
setwd("C:\\Users\\Star\\Documents\\my_data_files")# function to set the current working directory
```

```{r}
#| label: Load Dataset
#| echo: false
#| output: false
indicator_1 <- read.csv("unicef_indicator_1.csv")
indicator_2 <- read.csv("unicef_indicator_2.csv")
metadata <- read.csv("unicef_metadata.csv")

```

```{r}
#| echo: false
#| output: false
new_colnames <- c("country", "Alpha_Code2","Alpha_Code3" ,"Numeric_Code", "ISO3C", "Year", 
                  "Population_Total", "GDP_Constant_2015_USD", "GNI_Current_USD", 
                  "Inflation", "Life_Expectancy_at_Birth",
                  "Military_Expenditure_GDP_Perc")
names(metadata) <- new_colnames
```

```{r}
#| echo: false
#| output: false
colnames(metadata) # To check columns names
```

```{r}
#| echo: false
#| output: false
view(metadata)
```

```{r}
#| echo: false
#| output: false
# Remove rows with missing values
cleaned_metadata <- na.omit(metadata)
```

```{r}
#| echo: false
#| output: false
# Applying group by on data 
grouped_data <- cleaned_metadata %>%
  filter(Year == 2021) %>%
  group_by(country)
View(grouped_data)
```

## Bar Chart

The bar chart showcases the highest populated countries ranked in the order of China, India, the US, Indonesia, and Pakistan for 2021. China leads with the highest population due to its extensive landmass and historically rapid growth. India closely follows, also benefiting from a large geographical area and sustained population expansion. The US ranks third, leveraging economic prosperity to support its significant population. Indonesia and Pakistan, with expansive territories, secure their positions among the top populous nations

```{r}
#| echo: false
#| output: false
# Summarize the maximum population for each country
country_pop_summary <- summarise(grouped_data, max_population = max(Population_Total))
```

```{r}
#| echo: false
#| output: false
# Sort the summary by maximum population in descending order
country_pop_summary <- arrange(country_pop_summary, desc(max_population))
```

```{r}
#| echo: false
#| output: false
# Select the top 5 countries
top_5_countries <- head(country_pop_summary, 5)
```

```{r}
#| echo: false
#| output: true
# Create a data frame from top_5_countries (assuming top_5_countries is a character vector)
#top_5_countries_df <- data.frame(country = top_5_countries, max_population = top_5_countries)

# Create a bar chart
bar_chart <- ggplot(top_5_countries, aes(x = reorder(country, max_population), y = max_population)) +
  geom_bar(stat = "identity", fill = c("green","#97DFFC","yellow","blue","#DECDF5")) +
  labs(x = "Country", y = "Maximum Population", title = "Top 5 Countries by Maximum Population",
       subtitle = "By year 2021") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

# Display the bar chart
print(bar_chart)
 
```

## Time Series Plot

The time series plot vividly captures the population dynamics of China, India, the US, Indonesia, and Brazil spanning from 1960 to 2021. Notably, China experienced a staggering increase from 667,070,000 to 1,412,175,000, showcasing remarkable growth over the period. This massive rise underscores the imperative for robust Universal Health Coverage (UHC) frameworks to address the healthcare needs of expanding populations, ensuring equitable access to quality healthcare services despite demographic disparities.

```{r}
#| echo: false
#| output: false
#  Include data from the past 10 years (2012-2021)
filt_data <- cleaned_metadata %>%
  filter(Year >= 1960 & Year <= 2021)

```

```{r}
#| echo: false
#| output: false
# Group the data by country and summarize the population data for each year
grouped_data_TS <- filt_data %>%
  group_by(country, Year, .groups = "drop") %>%
  summarise(mean_population = mean(Population_Total))
```

```{r}
#| echo: false
#| output: false

top_5_countries <- grouped_data_TS %>%
  group_by(country) %>%
  summarise(avg_population = mean(mean_population)) %>%
  top_n(5, avg_population) %>%
  arrange(desc(avg_population)) %>%
  pull(country)

```

```{r}
#| echo: false
#| output: false
# Filter the grouped data to include only the top 5 countries
top_5_data <- grouped_data_TS %>%
  filter(country %in% top_5_countries)
```

```{r}
#| echo: false
#| output: false
#  Plot the time series for each of the top 5 countries
time_series_plot <- ggplot(top_5_data, aes(x = Year, y = mean_population, group = country, color = country)) +
  geom_line() +
  labs(x = "Year", y = "Mean Population", title = "Population Changes of Top 5 Countries (1960-2021)") +
  theme_minimal() +
  theme(legend.position = "right")  # To set Position of legend

```

```{r}
#| echo: false
#| output: true
# Display the time series plot
print(time_series_plot)
```

## Scatter plot with regression line:

The scatter plot depicts the relationship between life expectancy and population for Afghanistan, Albania, Algeria, Norway, and Pakistan, with life expectancy on the x-axis and population on the y-axis. Notably, Algeria's regression line appears steeper compared to Norway, suggesting a more pronounced positive impact on health systems and education in Algeria, leading to accelerated improvements in both life expectancy and population growth.

```{r}
#| echo: false
#| output: false
# Filter the data for countries "Afghanistan" ,"Norway" and "Pakistan"
filtered_metadata <- cleaned_metadata[cleaned_metadata$country %in% 
                                        c("Afghanistan", "Norway","Pakistan",
                                                                "Albania","Algeria",
                                                                      "Norway"), ]
```

```{r}
#| echo: false
#| output: false
# scatter plot with linear regression line
scatter_plot <- ggplot(filtered_metadata , aes(x = Year, y = Life_Expectancy_at_Birth, color = country)) +
  geom_point( size = 3, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # linear regression line
  labs(title = "Scatter Plot of Population vs. Life Expectancy",
       x = "Population",
       y = "Life Expectancy at Birth (Total Years)") +
  theme_minimal()

```

```{r}
#| echo: false
#| output: true
# Display the scatter plot
print(scatter_plot)
```

## World Map: Country Population for 2021

The world map illustrates population density across countries, with darker shades representing higher densities. This highlights areas where access to Universal Health Coverage (UHC) may be crucial due to concentrated populations. Factors such as job opportunities, living conditions, and government policies influencing density also impact the need for equitable healthcare provision. Understanding population density aids in planning UHC services effectively, ensuring all residents have access to essential healthcare regardless of population concentration.

```{r}
#| echo: false
#| output: false
# Load world map data
world_map <- map_data("world")

map_data <- grouped_data %>%
  left_join(world_map, by = c("country" = "region"), relationship = "many-to-many")

```

```{r}
#| echo: false
#| output: false
# Checking how the world map looks like
worldlayout <- ggplot() + 
  geom_polygon(data = world_map, aes(long, lat, group = group), fill = NA, color = 'grey') + 
  coord_fixed(1.3) + 
  theme_bw()

worldlayout
```

```{r}
#| echo: false
#| output: true
# Creating World Map  combined
map_Plot <- worldlayout + 
  geom_polygon(data = map_data, aes(x = long, y = lat, group = group, fill = Population_Total)) +
  scale_fill_viridis_b()+
  labs(title = "Country Population for 2021 ",
       x= "Logitude",
       y="Latitude")

map_Plot
```
